Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-05-27 doc/thesis/methods_results.tex
More of Sally's revisions

The accuracy of grid search estimates largely paralleled that of the \gls{kSVR}
classifiers. \Cref{fig:gridest} shows point estimates and 95\% highest density
intervals for each of the \gls{BA} parameters, for one replicate experiment
with 500-tip trees. Plots showing the point estimates for all replicates can be
found in \cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all
parameters except $m$, the error of point estimates was negatively correlated
with the number of sampled tips in the tree (for
\gls{alpha}, \gls{I}, and \gls{N} respectively: Spearman's $\rho$ = 
    \ensuremath{-0.22},
    \ensuremath{-0.51},
    \ensuremath{-0.16};
$p$-values
    $4\!\times\!10^{-4}$,
    ${<}10^{-5}$,
    $0.01$).
The highest density intervals obtained for all parameters were extremely wide,
occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).

The \gls{alpha} parameter was the most accurately estimated, with point
estimates having an average deviation of 
    0.14
from the true value, on a grid from 0 to 2. The error of point estimates varied
significantly between true values of \gls{alpha}
    (one-way \gls{ANOVA}, $p$ ${<}10^{-5}$). In
particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
for the other values
    (average errors 
    0.03
    for \gls{alpha} = 1.0 or 1.5 vs.
    0.17
    for \gls{alpha} $\neq$ 1.0 or 1.5),
and this difference was significant
    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
     \cref{fig:gridptalpha}).
These two values exhibited different qualitative behaviour than the other
values in terms of the distribution of kernel scores along the grid
(\cref{fig:gridalpha}). In particular, there was a pronounced peak in scores
around the true value, in contrast to the other values where the scores were
flat around the true value. The effect was most obvious for the value
\gls{alpha} = 1.25.

The average absolute error of the point estimates for \gls{I} was 
    310 individuals,
on a grid of 500 to 5000, and these errors differed between true values of
\gls{I}
    (one-way \gls{ANOVA}, $p =0.001$).
The errors for $2000 \leq I \leq 3000$ were higher than those for the other
values
    (average errors
     430
     for $2000 \leq I \leq 3000$ vs.
     250
     for $I < 2000$ or $I > 3000$),
and this difference was significant
    (Wilcoxon rank-sum test, $p =6\!\times\!10^{-4}$,
     \cref{fig:gridptI}).
Kernel score distributions for all test values exhibited a similar rounded
shape (\cref{fig:gridI}).

The average error for \gls{m} was
    1.31 edges per vertex,
on a grid from 1 to 6. The error varied significantly between the true values
of \gls{m} 
    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
Errors for the value \gls{m} = 1 were lower than the other values
    (average errors
    0.1
    for $m = 1$ vs.
    1.55
    for $m > 1$),
and this difference was significant
    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
     \cref{fig:gridptm}).
The value $m = 1$ causes the network to take on a distinct shape relative to
higher \gls{m} values, namely a tree (\ie there are no cycles, see
\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
when this was the true value, and a valley at $m = 1$ when the true value of
$m$ was greater that 1 (\cref{fig:gridm}).

The average error for \gls{N} was 
    2419 individuals,
on a grid from 1000 to 15000, and was varied significantly with the true value
of \gls{N}
    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
The errors were lower for $N \leq 3000$
    (average errors
     740
     for $N \leq 3000$ vs.
     2979
     for $N > 3000$),
and this difference was significant
    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
     \cref{fig:gridptN}).
The kernel score distribution had a peak at $N = 1000$ when this was the true
value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
the distributions were flat for $N > 3000$.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{gridsearch-example}
  \caption[Grid search estimates of \gls{BA} model parameters.]{Point estimates
      and 95\% highest density intervals for each \gls{BA} model parameter,
      obtained using grid search. Networks and transmission trees were
      simulated over a grid of values for each parameter while holding the
      others fixed. For a subset of the grid values ($x$-axis), test networks
      and trees were created and compared to each tree on the grid using the
      tree kernel. The kernel scores along the grid were normalized to resemble
      a probability distribution, from which the mode and highest density
      interval were calculated. Shown values correspond to one replicate
      experiment, with trees of size 500.
  } 
  \label{fig:gridest}
\end{figure}

@@

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
[95\% HPD] estimates for each parameter were: 
$\alpha$, 
  1.09 
  [0.6 -
   1.14];
$I$,
  1143 
  [666 -
   4455];
$m$,
  3 
  [2 -
   5];
$N$,
  12635 
  [4282-
   14977].
The approximate posterior distributions for this simulation are shown in
\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
transmission tree in a peer-driven fashion, where the probability to sample a
node was twice as high if one of its peers had already been sampled. The
parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
$I$ = 2000. The estimated values were
$\alpha$, 
  0.23 
  [0 -
   0.63];
$I$,
  2383 
  [1419 -
   3811];
$m$,
  4 
  [2 -
   5];
$N$,
  13517 
  [2836 -
   14720].
The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
of these results were in line with estimates obtained on other simulated
datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
$\alpha$ was somewhat lower than typical.

>>>>

{\color{red}\sout{The accuracy of grid search estimates largely paralleled that
of the \gls{kSVR} classifiers.}}\Cref{fig:gridest} shows point estimates and
95\% highest density intervals for each of the \gls{BA} parameters, for one
replicate experiment with 500-tip trees. Plots showing the point estimates for
all replicates can be found in
\cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all parameters
except $m$, the error of point estimates was negatively correlated with the
number of sampled tips in the tree (for \gls{alpha}, \gls{I}, and \gls{N}
respectively: Spearman's $\rho$ = \ensuremath{-0.22},
\ensuremath{-0.51}, \ensuremath{-0.16}; $p$-values
    $4\!\times\!10^{-4}$,
    ${<}10^{-5}$,
    $0.01$).
The 95\% highest density intervals obtained for all parameters were extremely
wide, occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).

{\color{blue}\uline{Similar to the results obtained in the previous experiment,
the \gls{alpha} and \gls{I} parameters were estimated with comparably high
accuracy, while the other two parameters were less identifiable. Across all
replicates, $R^2$ values for the correlations between the estimated and true
values were
    $0.91$,
    $0.91$,
    $0.25$, and
    $0.54$
for \gls{alpha}, \gls{I}, \gls{m}, and \gls{N} respectively. The mean absolute
errors of the point estimates were
    $0.14$,
    $310$,
    $1.31$, and
    $2419$,
representing
    $7$\%,
    $8$\%,
    $26$\%, and
    $17$\%
of the respective grids.}

\uline{Qualitatively, three of the parameters (excluding \gls{N}) exhibited
weak identifiability within particular sections of the grid. The \glspl{IQR}
for \gls{alpha} were nearly identical for the values \gls{alpha} $\leq 1$,
spanning approximately [0-1.3]. Similarly, the \glspl{IQR} for \gls{alpha} $>
1$ were all approximately [0.7-2]} (\cref{fig:gridest}). \uline{The
distributions of kernel scores for these groups of values were also
qualitatively similar} (\cref{fig:gridalpha}), \uline{although the value
\gls{alpha} = 1.25 exhibited a more pronounced peak in kernel scores around the
true value.}

\uline{For \gls{I}, the \glspl{IQR} were nearly identical for the values
\gls{I} $\leq 2000$ (approximately [500,3800]), and for the values \gls{I} $>
3000$ (approximately [1600-5000]). Kernel score distributions for all \gls{I}
values exibited a similar rounded shape} (\cref{fig:gridI}), \uline{although
this was most pronounced for values near the upper and lower extremes of the
grid.}

\uline{The \glspl{IQR} for \gls{m} were distinct for \gls{m} = 1 and 2, but
were similar for all values of \gls{m} $> 2$ (approximately [2-6]). The kernel
score distribution for \gls{m} = 1 was distinct from the others, having a large
peak at \gls{m} = 1, while the others exhibited a valley at \gls{m} = 1 and
were mostly flat everywhere else} (\cref{fig:gridm}).

\uline{The \glspl{IQR} for \gls{N} did not form any obvious groups like the
other parameters' \glspl{IQR}} (\cref{fig:gridest}). \uline{The kernel score
distribution for \gls{N} = 1000 was distinct, having a peak at the true value}
(\cref{fig:gridN}). \uline{The distributions for \gls{N} $> 1000$ were all
relatively similar, having a valley at the low end of the grid and being mostly 
flat everywhere else.}}

{\color{red}\sout{The \gls{alpha} parameter was the most accurately estimated,
with point estimates having an average deviation of 
    0.14
from the true value, on a grid from 0 to 2. The error of point estimates varied
significantly between true values of \gls{alpha}
    (one-way \gls{ANOVA}, $p$ ${<}10^{-5}$). In
particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
for the other values
    (average errors 
    0.03
    for \gls{alpha} = 1.0 or 1.5 vs.
    0.17
    for \gls{alpha} $\neq$ 1.0 or 1.5),
and this difference was significant
    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
}\cref{fig:gridptalpha}).
\sout{These two values exhibited different qualitative behaviour than the other
values in terms of the distribution of kernel scores along the grid
}(\cref{fig:gridalpha}). \sout{In particular, there was a pronounced peak in scores
around the true value, in contrast to the other values where the scores were
flat around the true value. The effect was most obvious for the value
\gls{alpha} = 1.25.}}

%The average absolute error of the point estimates for \gls{I} was 
%    d["I", round(mean(error))] individuals,
%on a grid of 500 to 5000, and these errors differed between true values of
%\gls{I}
%    (one-way \gls{ANOVA}, $p pp(av["I", p], eq=TRUE)$).
%The errors for $2000 \leq I \leq 3000$ were higher than those for the other
%values
%    (average errors
%     d["I", .SD[true_value >= 2000 & true_value <= 3000, round(mean(error))]]
%     for $2000 \leq I \leq 3000$ vs.
%     d["I", .SD[true_value < 2000 | true_value > 3000, round(mean(error))]]
%     for $I < 2000$ or $I > 3000$),
%and this difference was significant
%    (Wilcoxon rank-sum test, $p pp(I.test$p.value, eq=TRUE)$,
%     \cref{fig:gridptI}).
%Kernel score distributions for all test values exhibited a similar rounded
%shape (\cref{fig:gridI}). 
%
%The average error for \gls{m} was
%    d["m", round(mean(error), 2)] edges per vertex,
%on a grid from 1 to 6. The error varied significantly between the true values
%of \gls{m} 
%    (one-way \gls{ANOVA}, $p pp(av["m", p], eq=TRUE)$).
%Errors for the value \gls{m} = 1 were lower than the other values
%    (average errors
%    d["m", .SD[true_value == 1, round(mean(error), 2)]]
%    for $m = 1$ vs.
%    d["m", .SD[true_value > 1, round(mean(error), 2)]]
%    for $m > 1$),
%and this difference was significant
%    (Wilcoxon rank-sum test, $p pp(m.test$p.value, eq=TRUE)$,
%     \cref{fig:gridptm}).
%The value $m = 1$ causes the network to take on a distinct shape relative to
%higher \gls{m} values, namely a tree (\ie there are no cycles, see
%\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
%when this was the true value, and a valley at $m = 1$ when the true value of
%$m$ was greater that 1 (\cref{fig:gridm}).
%
%The average error for \gls{N} was 
%    d[parameter == "N", round(mean(error))] individuals,
%on a grid from 1000 to 15000, and was varied significantly with the true value
%of \gls{N}
%    (one-way \gls{ANOVA}, $p pp(av["N", p], eq=TRUE)$).
%The errors were lower for $N \leq 3000$
%    (average errors
%     d["N", .SD[true_value <= 3000, round(mean(error))]]
%     for $N \leq 3000$ vs.
%     d["N", .SD[true_value > 3000, round(mean(error))]]
%     for $N > 3000$),
%and this difference was significant
%    (Wilcoxon rank-sum test, $p pp(N.test$p.value, eq=TRUE)$,
%     \cref{fig:gridptN}).
%The kernel score distribution had a peak at $N = 1000$ when this was the true
%value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
%the distributions were flat for $N > 3000$.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{gridsearch-example}
  \caption[Grid search estimates of \gls{BA} model parameters.]{Point estimates
      (center lines) and 75\% (boxes) and 95\% (whiskers) highest density
      intervals for each \gls{BA} model parameter, obtained using grid search.
      Networks and transmission trees were simulated over a grid of values for
      each parameter while holding the others fixed. For a subset of the grid
      values ($x$-axis), test networks and trees were created and compared to
      each tree on the grid using the tree kernel. The kernel scores along the
      grid were normalized to resemble a probability distribution, from which
      the mode and highest density interval were calculated. Shown values
      correspond to one replicate experiment, with trees of size 500.
  } 
  \label{fig:gridest}
\end{figure}

@@

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
[95\% HPD] estimates for each parameter were: 
$\alpha$, 
  1.1 
  [0.6 -
   1.15];
$I$,
  1099 
  [662 -
   4455];
$m$,
  3 
  [1 -
   5];
$N$,
  12791 
  [3861-
   14967].
The approximate posterior distributions for this simulation are shown in
\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
transmission tree in a peer-driven fashion, where the probability to sample a
node was twice as high if one of its peers had already been sampled. The
parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
$I$ = 2000. The estimated values were
$\alpha$, 
  0.24 
  [0 -
   0.62];
$I$,
  2384 
  [1419 -
   3811];
$m$,
  3 
  [2 -
   5];
$N$,
  11565 
  [2741 -
   14780].
The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
of these results were in line with estimates obtained on other simulated
datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
$\alpha$ was somewhat lower than typical.
